---
title: "Income and Cardiovascular Health in LA County"
description: " Statical analysis of income and hypothesis testing on whether or not income can affect cardiovascular health "
author: 
  - name: Mariam Garcia
    url: https://mariamkg00.github.io/ 
    affiliation: MEDS
    affiliation-url: https://bren.ucsb.edu/masters-programs/master-environmental-data-science/academics-meds
date: 12-13-2023
image : los-angeles-2.png
format:
  html:
    code-fold: true
    code-summary: "Show the code"
categories: [Quarto, MEDS, Statistics, Environmental Justice]
citation:
  url: https://github.com/mariamkg00/eds-222-finalproject
draft: false
---

# Does Household Income Have an Affect on Cardiovascular Health?

## Motivation

According to the CDC, cardiovascular-related diseases are the leading cause of death in America. Although there are a number of factors that influence cardiovascular diseases, I chose to study the relationship between median income of census tracts in Los Angeles County and the number of cardiovascular disease-related Emergency Department (ED) visits. In this research project, I will conduct a difference in means hypothesis test of two groups : Extremely Low Income and Above Moderate Income to see if there is a true difference in heart attack-related Emergency Department (ED) visits.

Cardiovascular disease can be caused by a variety of factors. Some traditional risk factors include "male sex, older age, increased blood pressure, high cholesterol, low HDL (high density lipoprotein) and smoking"(epa.gov 2023). However, there are additional external factors that can exacerbate the likelihood of cardiovascular disease occurring. The EPA also states that people who live in areas with high air pollution are at an increased risk of PM2.5, along with people of low socioeconomic status.

Los Angeles is historically known to have strategically cut off and outline communities with freeways, which has led to brown and black communities predominantly having respiratory issues and other chronic health issues. According to the National Institute of Health, "COVID-19 has been reported to disproportionately impact racial/ethnic minorities in the USA, both in terms of infections and deaths" (Li, Yuan 2021). This can be explained by the historic segregation of minority groups. Neighborhoods that are currently undergoing socioeconomic disadvantages were known to be redlined historically. The New York case study concluded that the disproportionate number of deaths among minority groups in America during the COVID-19 pandemic highlighted the historic health and race divide compared to the White majority.

I chose this to research this topic because it is important to realize and note that environmental problems do not only include sea level rise or increased global temperature. Ongoing environmental problems combined with systemic racism can be detrimental to the survival of minority groups in America.

Furthermore, I decided to incorporate other variables in this research topic in order to explore other socioeconomic and chemical variables that could be correlated with the number of heart attack related emergency department visits. Education is an important socioeconomic variable to research as it can influence employment and income levels. According to the CDC , " individuals with lower levels of educational attainment are more likely to lack sociopolitical power and economic resources"(cdc.gov). Although not all socioeconomic factors are covered in this research project, education will play an important role in determining income level and housing status.

## Data

CalEnviroScreen has recently celebrated their tenth anniversary since releasing their first version. The CalEnviroScreen data set used in this study aims to highlight which communities within California are the most impacted by pollution by using census tract data that includes socioeconomic and demographic information which can help further encapsulate the larger issue at hand : pollution burden. Both California EPA and OEHHA (Office of Environmental Health Hazard Assessment) have collaboratively formulated this Environmental Justice screening tool as a means to help visualize impacts of pollution burden while accumulating various demographic data accessible for readers.

-   Variables Explored:

    -   `PM2.5` - Annual mean PM 2.5 concentrations (micrograms per meter cubed)

    -   `education` - Percent of population over 25 with less than a high school education

    -   `cardiovascular_disease` - Age-adjusted rate of emergency department visits for heart attacks per 10,000.

The second dataset used for this research project contains statistical information of median household income in LA County, which is then categorized into `ami categories` : Above Moderate Income, Moderate Income, Low Income, Very Low Income, and Extremely Low Income.

My game plan for analysis includes :

1.  Tidying both data sets
2.  Joining data sets by census tract
3.  Performing a linear regression
4.  Conducting a hypothesis test for a difference in means of emergency department visits for heart attacks between very low income AMI census tracts vs above moderate income AMI census tracts.

## Methods

```{r include = FALSE, message = FALSE , warning=FALSE}
# loading in needed libraries

library(tidyverse)
library(dplyr)
library(ggplot2)
library(stats)
library(lubridate)
library(janitor)
library(sf)
library(stargazer)
library(ggpmisc)
library(gt)
library(patchwork)
```

```{r message=FALSE}

# reading in the data 

income_la <- read_csv('./data/Median_Income_and_AMI_(census_tract).csv') %>% 
  rename('census_tract' = 'tract')

# reading in water la data

los_angeles <- read_csv('data/cal_enviro_4.csv') %>% 
  clean_names() %>% 
  filter(california_county == 'Los Angeles')


# joining in the two datasets

income_water_la <- left_join(los_angeles, income_la, by = "census_tract") %>% 
  na.omit()


# preliminary plotting
income_cardio_tox_ed <- lm(cardiovascular_disease ~ med_hh_income + tox_release +education, data = income_water_la)

income_cardio <- lm(cardiovascular_disease ~ med_hh_income, data = income_water_la)

summary(income_cardio)




```

```{r}
count_income <- income_water_la %>% 
  group_by(ami_category) %>% 
  count() %>%  
  rename(`Number of Census Groups` = n,
         `AMI Category` = ami_category)
knitr::kable(count_income)

```

Of the 1,905 census tract groups with an associated AMI category :

-   419 census tract groups fall under the `Above Moderate Income` category

-   633 census tract groups fall under the `Very Low Income` category.

```{r warning = FALSE, message= FALSE}
ggplot(income_water_la, aes(x = med_hh_income, y = cardiovascular_disease, col = education))+
  geom_point()+
  stat_smooth(method = 'lm', col = 'red')+
  labs(x = 'Median Household Income (USD)',
       y = 'Heart Attack ED visits per 10,000 people',
       title = '     Median Income vs. Cardiovascular Disease')+
  theme_classic()+
  stat_poly_eq()
```

```{r}
ggplot(income_water_la, aes(x = education, y = cardiovascular_disease))+
  geom_point(color = '#CBC3E3')+
  stat_smooth(method = 'lm', col = '#9F2B68')+
  labs(x = '% of Adults > 25 with Less \nThan a High School Education',
       y = 'Heart Attack ED visits per 10,000 people',
       title = '     Education Level vs. Cardiovascular Disease')+
  theme_classic()+
  stat_poly_eq()
```

```{r}
## hypothesis test construction
# summary statistics for cardiovascular disease related ed visits

cardio_stats <- income_water_la %>% 
  group_by(ami_category) %>% 
  summarise(mean = mean(cardiovascular_disease),
            sd = sd(cardiovascular_disease))


print(cardio_stats)
```

The `Extremely Low Income` AMI group has the most variance amongst AMI groups in this dataset.

However, the number of census tract groups in the `Extremely Low Income` AMI group is significantly lower than the number of census groups in the `Above Moderate Income`. In an effort to make this analysis as representative as possible, I will be using the data from census tract groups in the `Very Low Income` AMI category.

Using the `stargazer` package, I will conduct three different linear models to determine which added variables positively affect the adjusted **R2** value.

```{r warning = FALSE}
# table of r squared values with different variables added 

m3 <- lm(cardiovascular_disease ~ med_hh_income + pm2_5 + education, data = income_water_la)
m2 <- lm(cardiovascular_disease ~  pm2_5, data = income_water_la)
m1 <- lm(cardiovascular_disease ~ med_hh_income, data = income_water_la)


stargazer(m1, m2, m3, title = 'Regression Results',type="text", dep.var.labels = c('Heart Attack Related ED Visits per 10,000'),
          covariate.labels=c("Median Household Income","PM 2.5 Concentration","% Population > 25 with less than a High School Education"),
          omit.stat = c('f'),
          font.size = 'small',align = TRUE)
```

**Interpretation of stargazer table:**

[Output for first equation `m1`:]{.underline}

-   For every 1 unit increase in `Median Household Income`, the rate of cardiovascular disease related ED visits is expected to *decrease* by -0.0001 per 10,000.

    -   To put this in perspective, a 10,000 increase in `Median Household Income` is expected to decrease the rate of ED visits for heart attacks by 1 per 10,000.

-   **R2** : 0.226 of all variability in the `cardiovascular_disease` output can be explained by `Median Household Income`.

[Output for second equation `m2`]{.underline} :

-   A one unit increase in annual particulate matter concentration `pm2_5` is expected to increase ED visits for heart attacks by .254 per 10,000.

-   **R2** : .002 of all variability in the `cardiovascular_disease` output can be explained by `pm2_5`.

[Output for third equation `m3`]{.underline} :

-   Holding `pm2_5` and `Median Household Income` constant, a one percent increase in population older than 25 with less than a high school education will lead to a 0.119 per 10,000 increase in ED visits for heart attacks.

-   The adjusted R2 value is 0.319, meaning that , combined, `Median Household Income`, `Particulate Matter Concentration`, and `Education` explain 31.9% of variability in the rate of emergency department visits for heart attacks per 10,000.

```         
```

$$H_{0}: \mu_{cardiodiseaseinVeryrLowInc} - \mu_{cardiodiseaseinModInc}  = 0$$ $$H_{A}:  \mu_{cardiodiseaseinVeryLowInc} - \mu_{cardiodiseaseinModInc} \neq 0$$

$$SE = \sqrt{\frac{s_1^2}{n_1} + \frac{s^2_2}{n_2}}$$

```{r}

# computing a point estimate for both groups
mu_cardio_low <- income_water_la %>% 
  filter(ami_category == 'Very Low Income') %>% 
  summarize(mean(cardiovascular_disease))

mu_cardio_mod <- income_water_la %>% 
  filter(ami_category == 'Above Moderate Income') %>% 
  summarize(mean(cardiovascular_disease))


point_est_cardio <- as.numeric(mu_cardio_low - mu_cardio_mod) # will use for z-score 

print(point_est_cardio)


```

The point estimate value of `5.98` indicates that there is a difference between the two samples.

```{r}

# computing standard error

n1 <- income_water_la %>% 
  filter(ami_category == 'Very Low Income') %>% 
  count()

n2 <- income_water_la %>% 
  filter(ami_category == 'Above Moderate Income') %>% 
  count()

s1 <- income_water_la %>% 
  filter(ami_category == 'Very Low Income') %>% 
  summarize(sd(cardiovascular_disease, na.rm = TRUE))

s2<- income_water_la %>% 
  filter(ami_category == 'Above Moderate Income') %>% 
  summarize(sd(cardiovascular_disease, na.rm = TRUE))


# computing standard error

standard_error <-as.numeric(sqrt(s1^2/n1 + s2^2/n2))


# computing z-score
z <- (point_est_cardio - 0)/standard_error


```

The z-score of `24.22` indicates that the observed difference in cardiovascular disease related ED visits between Above Moderate Income census tracts and Very Low Income census tract groups is 24.2 standard deviations **above** the null of zero difference.

Now that I have a calculated z-score, I am then able to calculate the p-value. I decided to input `lower.tail = TRUE` because I am finding the probability that the lower tail of my confidence interval is no bigger than my z score of `24.22`.

\* In other words, I am using `lower.tail = TRUE` because we are trying to calculate the probability to the **left** of 0 .

```{r}
p_value <- pnorm(z,lower.tail = FALSE)
print(p_value)
```

Alternatively, this hypothesis test can be executed using a `t.test()`

```{r}
t.test(income_water_la$cardiovascular_disease[income_water_la$ami_category == 'Very Low Income'], income_water_la$cardiovascular_disease[income_water_la$ami_category == 'Above Moderate Income'])
```

Based on the output by the `Welch Two Sample t-test`, I am 95% confident that (5.49, 6.46) captures the true value of the average difference of cardiovascular disease related ED visits between the two income groups.

The p-value derived from the `Welch Two Sample t-test` is significantly higher than the p-value derived from the `pnorm` calculation, and this could be explained by the fact that the `Welch Two Sample t-test` assumes it is a t-distribution.

##### Calculating a 95% confidence interval

I'd like to construct a confidence interval apart from the derived interval in order to calculate a 5% probability that I am able to observe the point estimate. Calculating this confidence interval means that I will likely be 5% wrong that the difference between ED visits for heart attacks is between this range.

```{r}
# 95% confidence interval 

critical_val <- qnorm(0.025, lower.tail = FALSE)

lower_ci <- round(point_est_cardio - critical_val*standard_error, 2)
upper_ci <- round(point_est_cardio + critical_val*standard_error,2)

print(paste0("There is a 95% probability that [",lower_ci, ",", upper_ci, "] contains the difference in cardiovascular related ED visits across above moderate income and extremely low income census tract groups. "))
```

## Results and Discussion

Understanding the impact that socioeconomic status can have on a census tracts' overall health is important in order to address the larger issue at hand : environmental racism. Through hypothesis testing and linear regression, I was able to conclude that cardiovascular health does depend on a variety of factors that are outside the scope of the data sets used, but I was able to point out the disparities between two vastly different income groups.

Based on the `stargazer` results, adding more variables to the third linear model did increase the `R2` value, but not enough to conclude that there is a strong correlation as its value was less than 0.5. The `p-value of 6.561525e-130` indicates that it is **very unlikely** to see the point estimate if the null hypothesis were true. In other words, there is a 5% chance that observing the point estimate of 5.98 would occur even if the null hypothesis were true.

This is consistent with the research provided by the EPA stating that the occurrence of cardiovascular disease does rely on a number of factors. I decided to use income as the explanatory variable because ,generally speaking, households in LA County that are of low income tend to live closer to freeways or polluting facilities.

Both methods of conducting the hypothesis test of the difference in two means calculated a p-value of less than 0.05. Thus, we can **reject the null** that there is no difference in heart attack related ED visits across above moderate income and very low income census tract groups. In conclusion, I found a statistically significant effect in which average income lowers the heart attack related ED rate per 10,000 people.

In the future, consistent units for certain variables like `drinking_water` would be helpful in running a deeper analysis. As the unit for this variable is the sum of percentiles of selected contaminants, it can be challenging to associate unit increases when percentiles come into play.

## Conclusion

Although income level isn't the only explanatory variable for cardiovascular health, this study highlighted the other potential socioeconomic factors that can be affected by income. Further research on this topic needs to be conducted, and possibly done in other states to confirm whether or not the effect is consistent in other parts of the world. Moreover, this analysis does prove there to be a correlation and statistically significant significance between income and cardiovascular health.

## References

*Air Pollution and Cardiovascular Disease Basics \| US EPA*, www.epa.gov/air-research/air-pollution-and-cardiovascular-disease-basics.

Li, Min, and Faxi Yuan. "Historical Redlining and Resident Exposure to Covid-19: A Study of New York City." *Race and Social Problems*, U.S. National Library of Medicine, 2022, www.ncbi.nlm.nih.gov/pmc/articles/PMC8212581/#CR3.

"Socioeconomic Factors." *Centers for Disease Control and Prevention*, 1 Sept. 2023, www.cdc.gov/dhdsp/health_equity/socioeconomic.htm#:\~:text=Socioeconomic%20status%20is%20a%20multidimensional,employment%20status%2C%20and%20other%20factors.&text=Low%20socioeconomic%20status%20is%20associated,from%20cardiovascular%20disease%20(CVD).

## Data Access

CalEnviroScreen Data : https://oehha.ca.gov/calenviroscreen/report/calenviroscreen-40

Median Income and AMI (census tract) : https://egis-lacounty.hub.arcgis.com/datasets/lacounty::median-income-and-ami-census-tract/about

### Supplemental Figures

```{r}
# qqplot
ggplot(data = income_water_la)+
  geom_qq(aes(sample = income_water_la$cardiovascular_disease), color = '#352C9E')+
  geom_qq_line(aes(sample = income_water_la$cardiovascular_disease))+
  theme(line = element_blank())+
  labs(x = 'Normal Distribution Quantiles',
       y = 'Sample Quantiles',
       title = 'Heart Attack-Related ED Visits (per 10,000)')+
  theme_classic()
```

This `qqplot` of the cardiovascular disease variable tells me that there is a slight divergence from the normal distribution line in the lower quantiles, indicating that values in this quantile range are higher relative to a normal distribution. Moreover, this `qqplot` also tells me that it follows a normal distribution in the higher quantiles.

```{r}

ggplot(data = income_water_la, aes(x = med_hh_income))+
  geom_histogram(bins = 100, fill = '#AAACE4')+
  labs(x = 'Median Household Income (USD)',
       y = 'Number of Census Tract Groups',
       title = 'Distribution of Median Income of LA County Census Tract Groups')+
  theme_minimal()
```
